---
title: Gene ontology enrichment of TWAS clusters
author: "Yongjin Park"
theme: jekyll-theme-minimal
date: "`r Sys.time()`"
bibliography: gtex.bib
csl: nature.csl
---

```{r GlobOpt, include = FALSE}
dir.create('Fig/20180808/', recursive = TRUE)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.path = 'Fig/20180808/')
```

```{r ReadData}
source('../Util.R')
source('./Util.Fig.R')
library(dplyr)
library(tidyr)
library(qvalue)
library(readr)

################################################################
twas.data.file <- '20180807_twas_result.RData'
load(twas.data.file)

required.twas.clust.file <- '../twas/hsb_7_0.05_cluster.RData'
load(required.twas.clust.file)

twas.ordering.data.file <- '20180807_twas_clustering.RData'
load(twas.ordering.data.file)

num.int <- function(x) format(x, big.mark = ',')
num.sci <- function(x, d=2) format(x, scientific = TRUE, digits = d)
num.round <- function(x, d=2) round(x, d)
gg.save <- function(...) ggsave(..., useDingbats = FALSE, units = 'in', limitsize = FALSE)
```

```{r TestEnrichment}
## Summary of enrichment within cluster
summary.enrichment.stat <- function(tab) {
    ## What can we control beyond the set size?

    ## calculate hypergeometric p-value
    ## phyper(q, m, n, k, lower.tail = TRUE, log.p = FALSE)
    ##    q: vector of quantiles representing the number of white balls
    ##       (drawn without replacement from an urn which contains both
    ##       black and white balls).
    ##    m: the number of white balls in the urn.
    ##    n: the number of black balls in the urn.
    ##    k: the number of balls drawn from the urn.
    ##    p: probability, it must be between 0 and 1.

    m <- max(tab$white.ball) ## white balls
    n <- max(tab$ntot) - m   ## black balls
    k <- length(tab$drawn)   ## balls drawn
    q <- sum(tab$drawn)      ## whilte balls drawn

    pval <- phyper(q, m, n, k, lower.tail = FALSE)

    data.frame(mu = mean(tab$drawn),
               n.white = m,
               n.drawn = k,
               n.overlap = q,
               pval = pval)
}

genes.tot.heritable <- fqtl.snp.tab %>% select(ensg) %>%
    unique() %>%
    separate('ensg', c('ensGene', 'remove'), sep = '[.]') %>%
    select(-remove) %>%
    .unlist()

temp.data.file <- '20180808_twas_clustering_geneontology.RData'

if(file.exists(temp.data.file)) {
    load(temp.data.file)
} else {

    ensembl <- biomaRt::useEnsembl(biomart='ensembl',
                          dataset = 'hsapiens_gene_ensembl',
                          version = 79)

    ## go.attr <- c('ensembl_gene_id', 'goslim_goa_accession', 'goslim_goa_description')

    go.attr <- c('ensembl_gene_id', 'go_id', 'name_1006', 'namespace_1003')

    go.bm <- biomaRt::getBM(filters = 'ensembl_gene_id', values = genes.tot.heritable,
                   attributes = go.attr,
                   mart = ensembl)

    go.tab <- go.bm %>% (function(tab) {
        names(tab) <- c('gene', 'go', 'go.name', 'go.onto');
        return(tab) }) %>%
            filter(nchar(go.name) > 0)

    go.size.tab <- go.tab %>%
        group_by(go) %>%
            summarize(go.size = n()) %>%
                as.data.frame()

    ntot <- go.tab %>% dplyr::select(gene) %>%
        unique() %>% nrow()
   
    clust.out <- twas.factor.cluster

    gene.cluster <- 
        clust.out$Z %>%
            tidyr::separate('ensg', c('gene', 'remove'), sep = '[.]') %>%
                dplyr::select(-remove, -factor) %>%
                    tidyr::gather(key = 'cluster', value = 'val', -gene) %>%
                        mutate(cluster = gsub(cluster, pattern = 'K', replacement = '')) %>%
                            mutate(cluster = as.integer(cluster)) %>%
                                mutate(val = round(val))

    cluster.gene.ontology.tab <-
        gene.cluster %>%
            left_join(go.tab, by = 'gene') %>%
                left_join(go.size.tab, by = 'go') %>%
                    mutate(ntot = ntot) %>%
                        filter(go.size >= 5, go.size <= 500) %>%
                            group_by(cluster, go, go.name, go.onto) %>%
                                dplyr::rename(drawn = val, white.ball = go.size) %>%
                                    dplyr::do(summary.enrichment.stat(.)) %>%
                                        as.data.frame()

    save(list = c('cluster.gene.ontology.tab', 'gene.cluster'),
         file = temp.data.file)
}
```

We tested significant over-representation of the gene ontology (GO) terms [@Ashburner2000-nu; @The_Gene_Ontology_Consortium2017-wz] within each trait cluster, using `r genes.tot.heritable %>% length() %>% num.int()` all the heritable genes identified in this work as a background set.  We downloaded mapping between genes and GO terms using `biomaRt` and its `R` interface [@Durinck2005-fk; @Durinck2009-tu] (ENSEMBL version 79).  We restricted our tests on `r cluster.gene.ontology.tab$go.name %>% unique() %>% length() %>% num.int()` GO terms with a reasonable set size, ranging between 5 and 500 after intersecting them with the heritable genes.

```{r Fig_Top_GO_clust, fig.width = 10, fig.height = 8, fig.cap = '\\label{fig:go.tab}__Gene Ontology enrichement on TWAS clusters shows disnticitive biological processes.__'}
################################################################
cluster.gene.ontology.tab <- 
    cluster.gene.ontology.tab %>%
    mutate(argmax = factor(cluster, trait.clust.order)) %>%
    na.omit()

take.top.go <- function(tab) {
    ret <- tab %>%
        filter(n.overlap > 0, mu > .1) %>%
            arrange(pval, mu, n.white) %>%
                head(3)
    ret %>% as.data.frame()
}

top.clust.go <- cluster.gene.ontology.tab %>% 
    filter(n.overlap > 0, n.drawn > 2, mu > .1) %>%
    filter(go.onto == 'biological_process') %>%
    filter(pval < 1e-5) %>%
    group_by(cluster) %>%
    do(take.top.go(.)) %>%
    as.data.frame()

go.clust.show <- 
    top.clust.go %>% select(go.name) %>% unique() %>%
    left_join(cluster.gene.ontology.tab)
    
go.names.show <- go.clust.show$go.name %>% unique() %>% sort()

go.name.order <-
    go.clust.show %>%
    mutate(argmax.i = as.integer(argmax)) %>%
    select(go.name, argmax.i, pval) %>%
    spread(key = argmax.i, value = pval) %>%
    as.data.frame() %>%
    arrange(go.name) %>%
    select(-go.name) %>%
    apply(MARGIN = 1, FUN = which.min) %>%
    order(decreasing = TRUE) %>%
    ((function(x) go.names.show[x] ))

go.clust.show.sorted <- go.clust.show %>%
    mutate(go.name = factor(go.name, go.name.order)) %>%
    mutate(nl.p = pmin(-log10(pval), 8))

.go.fill.scale <- scale_fill_gradientn('-log10\np-value',
                                       colors = c('white', 'white', 'orange', 'red'))

.go.df.show <- go.clust.show.sorted %>%
    filter(pval < 1e-4) %>%
    mutate(lab = argmax)

plt.go.clust <- 
    gg.plot(go.clust.show.sorted) +
    geom_tile(aes(x = argmax, y = go.name, fill = nl.p), color = 'gray') +
    geom_text(data = .go.df.show, 
              aes(x = argmax, y = go.name, label = lab),
              size = 1.5)

plt.go.clust <- plt.go.clust +
    .go.fill.scale +
    theme(axis.title = element_blank(),
          axis.text.y = element_text(size = 6.5),
          axis.text.x = element_text(size = 4)) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

print(plt.go.clust)
gg.save(filename = 'Fig/20180808/Fig_Top_GO_clust.pdf', plot = plt.go.clust, width = 10, height = 8)
```

Of the tested `r cluster.gene.ontology.tab %>% dplyr::select(go) %>% unique() %>% nrow() %>% num.int()`, total `r sum(p.adjust(cluster.gene.ontology.tab$pval) < .05) %>% num.int()` GO terms are significantly over-represented with family-wise error rate cutoff 0.05, and we highlight the `r go.clust.show.sorted$go.name %>% unique() %>% length()` terms in `biological process` with nominal p-value $<$ 1e-5, sampling at most three most enriched categories from within each cluster (Fig.\ref{fig:go.tab}).


# References
